### 2022-10-03 biblio-rp.tex
initial commit

\documentclass[11pt, twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{fdsymbol}
\usepackage{textgreek}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{url}
\usepackage{array}
\usepackage{csquotes}
\usepackage{caption}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}

\title{Bibliographic Survey\\\medskip Research Project 2022}
\author{Nicolas COURAUD (nicolas.couraud@etu.emse.fr)\\École Nationale Supérieure des Mines de Saint-Étienne}

\begin{document}

\maketitle
\onecolumn
\section*{Introduction: Why Consensus ?}

The boom of cryptocurrency that the world of tech/finance has seen in the past years (2017, 2021) shed light on a domain of computer science which is in fact much older: consensus algorithms. While the Bitcoin whitepaper was released in 2008 and the first block of the bitcoin blockchain was mined in early 2009, it is after around a decade of nearly flawless functioning of the bitcoin blockchain that institutions started looking at it, and investments started flowing in.
Bitcoin introduced a new consensus algorithm, based on Proof-of-Work, an energy-consuming mechanism designed to resist attackers.

Nonetheless, the research on consensus algorithms goes back way before the internet even existed.
So, why is consensus needed in computed science ?

\tableofcontents
\section{The need for consensus}

\subsection{Outside computer science}

In order to better understand the world of consensus algorithms in computer science, we can take an example from an early democratic systems in Greece. Let's imagine that the law of a country is defined by a unique series of decrees that are passed with a majority of voters in a parliament \cite{parliament} . The parliament must function even though legislators constantly go in and out of the parliament.
No single record of the law can exist, because those can be corrupted, and it would require one person to be in charge of the transcribing the votes all day long.

The parliament therefore requires a consensus protocol to make sure that legislators don't have contradictory information on there version of the law (e.g. 2 different decrees with the same number).

You might think that a majority vote is all what's required to establish new decrees, but some questions on the protocol quickly come to mind:
\begin{itemize}
    \item Who is allowed to propose new decrees, and when ?
    \item How to make sure the consensus is legit when voters are wandering in and out of the parliament ?
    \item When a new parliamentary want the current state of law, where can he get it, or from whom ?
\end{itemize}

All those interrogations are discussed in the research on consensus algorithms.

\subsection{In computer science}

Let’s imagine that one need to run a server to access online versions of self-published books. That is fairly easy, you need a computer that handles requests from clients, processes them through a search engine and then answers with a list of books corresponding to the search.

Now as soon as you need to scale the server, you will need the database to be distributed among multiple computers or servers.
Then, what happens if a user, author of one of the book, releases a new version at tries to update the online version hosted on the server ?
How can one make sure that the version will be available on all instances ? If we choose to have one “leader” from which all the servers copy the state of the data, how is that leader chosen by the protocol, and changed if that node crashes ?

When there is decentralised processes, there are consensus algorithms behind, such as in etcd \cite{etcd}, AWS\footnote{Amazon Web Services} and cloud computing in general \cite{aws}, bitcoin and blockchain \cite{bitcoin}.

\section{Evaluation consensus protocols}
\subsection{Defining the consensus}

As we stated above, distributed consensus is the problem of getting multiple entities (servers) to agree on a shared state, even in the case of failures.
That shared state is usually some data structure backed by a replicated log.
The property we want from a consensus system is that it is fully operational, even if a minority of the servers fail.
If the majority of servers are up, system is up.

Consensus among a set of decentralized nodes can have multiple definitions.
We will review some of the definitions that are available in the the scientific literature before
we study any consensus algorithm to have a better understanding of the range of situations these algorithms can apply to.
Most consensus definitions have variants.

\paragraph{Consensus on one value}

The simplest form of consensus is the consensus on one bit ( one boolean ) and is defined as follows by Michael J. Fischer \cite{consensusproblem}:

\begin{displayquote}
    Assume a fixed number of processors, some of which are initially faulty or may fail during the execution of the protocol.
    Each processor $i$ has an initial bit $x_i$. The consensus problem is for the non-faulty processes to agree on a bit $y$, called the consensus value.
    More precisely, we want a protocol such that each reliable process $i$ eventually terminates with a bit $y_i$, and \( \forall i, y_i == y \). $y$ in general will depend in some way on the initial bits $x_i$.
    In the absence of such a requirement, the problem becomes trivial, for each process can simply choose $y_i = 0$.
    Some dependency requirements that have been studied, in order of increasing strength, are:
    \begin{enumerate}
        \item (non-triviality): For each \( y \in (0, 1)\), there is some initial vector \( x_{i}\) and some admissible execution of the protocol in which $y$ is the consensus value. (The qualification "admissible" allows for additional restrictions, such as bounds on the number of faulty processes, on the kinds of computations we are willing to consider)
        \item (weak unanimity): If \(\forall i,\;x_{i} = x \in (0, 1)\), then \(y == x\), provided that no failures actually occur during the execution of the protocol.
        \item (strong unanimity): If \(\forall i,\;x_{i} = x \in (0, 1)\), then \(y == x\).
    \end{enumerate}

\end{displayquote}
This definition is simple to understand but very restrictive.

\paragraph{A better definition}

Michel Raynal gives us a broader definition of the consensus problem in his book \cite{raynal} :
\begin{displayquote}
    In the consensus problem, each process $p_i$ proposes a value $v_i$ and all processes have to agree (we also say “decide”) on the same value that has to be one of the proposed values.

    In a more precise way, this problem is defined by the following properties :
    \begin{itemize}
        \item Validity : If a process decides a value $v$, then $v$ has been proposed by some process.
        \item Integrity : A process decides at most once.
        \item Agreement : No two processes decide different values.
        \item Termination : Each correct process decides.
    \end{itemize}
    Let A be an algorithm that is assumed to solve the consensus problem. This algorithm is correct if all its runs satisfy the previous properties. Validity, integrity and agreement define the consensus safety properties, while termination defines its liveness property.
\end{displayquote}

As in others studies of consensus problems, Raynal highlights that the consensus can be solved in a trivial way in failure-free asynchronous systems, for example agreeing on the value decided by the process with the smallest number (assuming processes are numbered).

\paragraph{Interactive consistency}

Interactive consistency is a notion defined by M. Pease, R, Shostak, and L. Lamport in 1980 \cite{reachingagreement}.

\begin{displayquote}
    Consider a set of n isolated processors, of which it is known that no more than m are faulty.
    It is not known, however, which processors are faulty. Suppose that the processors can communicate only by means of two-party messages.
    The communication medium is presumed to be fail-safe and of negligible delay. The sender of a message, moreover, is always identifiable by the receiver.
    Suppose also that each processor $p$ has some private value of information \(V_{p}\). [...]
    The question is whether for given \(m, n \geqslant 0 \), it is possible to devise an algorithm based on an exchange of messages that will allow each nonfaulty processor $p$ to compute a vector of values with an element for each of the n processors, such that :

    \begin{itemize}
        \item the nonfaulty processors compute exactly the same vector;
        \item the element of this vector corresponding to a given nonfaulty processor is the private value of that processor.
    \end{itemize}

    Note that the algorithm need not reveal which processors are faulty, and that the elements of the computed vector corresponding to faulty processors may be arbitrary; it matters only that the nonfaulty processors compute exactly the same value for any given faulty processor.
    We say that such an algorithm achieves interactive consistency, since it allows the nonfaulty processors to come to a consistent view of the values held by all the processors, including the faulty ones. The computed vector is called an interactive consistency (i.c.) vector. Once interactive consistency has been achieved, each nonfaulty processor can apply an averaging or filtering function to the i.c. vector, according to the needs of the application.
    Since each nonfaulty processor applies this function to the same vector of values, an exact agreement is necessarily reached.

\end{displayquote}

Interactive consistency differs from consensus because instead of having the nodes agreeing on one common value, we want nodes agreeing on a vector which size corresponds to the number of nodes in the system.

\paragraph{Consensus stability}

The consensus stability clause is a clause to define the type of consensus. Different levels of consensus stability can be required depending on our problem.

The three main consensus stability values are, from the strongest consensus to the weakest :

\begin{itemize}
    \item Irrevocable: Each agent has only one output value during the whole consensus process.
    \item Stabilization: Each agent's output values converge in finite time : there is a final stable value.
    \item Asymptotic : Each agent's output values converge.
\end{itemize}

Nakamoto consensus\footnote{\nameref{sec:pow} (\ref{sec:pow})} used in the bitcoin blockchain lands between irrevocable and stabilizing consensus \cite{proconsensus}.

When studying irrevocable consensus, we have to keep in mind an important theorem called FLP (Fischer, Lynch, Paterson) \cite{fischerlp}:
In an asynchronous system of processes, every protocol trying to solve the consensus has the possibility of non termination, even with only one faulty process.
Another way of stating FLP is to say that in an asynchronous system, there is no irrevocable consensus algorithm that tolerates one crash.

\paragraph{Uniform consensus}

The uniform consensus is introduced \cite{uniformConsensus} to tackle a specific case of consensus reach, which fits the consensus definition, but could still lead to system having a unwanted / unsustainable behaviour.
In the case of a failure occurring on a process a long time after it was solicited, because the traditional consensus definition ignores faulty processes (and a faulty process could show a failure after being solicited), two processes could have reached consensus on two different values, one of the processes fail, and this way consensus is reached.

Indeed, the uniform consensus adds up a new condition (called uniform agreement) to the consensus and thereby strengthens the agreement:
No two processes (whether faulty or not) can decide differently.

In synchronous systems, in the crash-failure model (no byzantine failure\footnote{\nameref{subsec:byzantine} (\ref{subsec:byzantine})}), if the uniform consensus is required, one additional round is needed to decide, so uniform consensus is harder than regular consensus for crash failures.

\subsection{Consensus and synchrony}
\paragraph{Synchrony and asynchrony}

Consensus algorithms and results will differ a lot between a synchronous and an asynchronous system. Let's define those terms :
A synchronous system of processes has :
\begin{itemize}
    \item message delays < Δ
    \item relative computing speed < Φ
    \item Δ and Φ are known
\end{itemize}

Those are the hypothesis that correspond to synchrony.
Asynchronous systems are all systems not bound by those hypothesis

Assuming the system is synchronous changes the impossibility theorem for irrevocable consensus.
While in an asynchronous system, the FLP result restricts most of the work on consensus, the impossibility result in synchronous system is a bit less strict on what cannot be done.

That result is called SPL, standing for Shostak, Pease \& Lamport and dates back to 1982 \cite{byzantine}. It states the following :
In a synchronous system, there is no irrevocable consensus algorithm that tolerates \(n/3\) byzantine agents.

We will discuss in the next section what a byzantine agent is.

\paragraph{Partial synchrony}

Finally, the last result worth mentioning states a possibility and not an impossibility in partially synchronous systems \cite{partialSynchrony}.

Partially synchronous systems are systems laying between synchronous and asynchronous systems. One example would be a system in which messages delays and relative computing speed are bound, but we don't know the value of the bounds a priori. In another case, bounds might be known, but only apply starting as some unknown time T, whereas the system must work correctly at any time.

The result demonstrated by Dwork, Lynch \& Stockmeyer (DLS) is the following :
For partially synchronous systems, there is an irrevocable consensus algorithm that tolerates a minority of crashes.

The “minority” value depends on the type of crashes that the system is facing, the number of processes affected by those crashes and the type of partial synchrony that constitutes our hypothesis. The DLS paper gives the values for each situation of how many processes can be faulty for the whole system to stay resilient.

\subsection{Fault Tolerance}

Let’s dive deeper in what we mean by fault tolerance in consensus algorithms.
We already stated that having a consensus among a set of processes, of which none is faulty is trivial. We are now going to differentiate failure types in order to be able to assess different types of consensus algorithms.

Failures definitions come from \cite{partialSynchrony}. The definitions take place in a system defined as follows :
\begin{displayquote}
    The communication system is modeled as a collection of N sets of messages, called buffers, one for each processor. The buffer of pi represents messages that have been sent to $p_i$ , but not yet received. Each processor follows a deterministic protocol involving the receipt and sending of messages. Each processor $p_i$ can perform one of the following instructions in each step of its protocol:
    \(Send(m, p_j)\): places message m in $p_j$'s buffer
    \(Receive(p_i)\): removes some (possibly empty) set S of messages from $p_j$'s buffer and delivers the messages to $p_i$ .
\end{displayquote}

\paragraph{Crash failures}

The crash failure is the most basic type of crash, and one of the most common. The process basically stops running. Therefore, it is unable to handle any requests for an undetermined period of time. Crash failures can occur in real life when the internet is down on a server, when there is a black out in the area (no more electricity), or if there is a maintenance for example.

Recently, the OVHcloud datacenter fire incident reminded the public that crashes can occur any time and anywhere.

The formal hypothesis for a processor $p_i$ with crash failure is : processor $p_i$ executes correctly, but can stop at any time. Once crashed it cannot restart

\paragraph{Omission failures}

The omission failure is considered a higher level of failure severity compared to the crash.
A process $p_i$ is subject to omission failure when :
\begin{itemize}
    \item \(Send(m, p_j)\) by \(p_i\) might not place $m$ in \(p_j\)'s buffer.
    \item \(Receive(p_i)\) might cause only a subset of the delivered messages to be actually received by \(p_i\).
\end{itemize}

Crash failures are sometimes referred to as a special case of omission failures, where the system can detect with certainty if that the process has crashed.

\paragraph{Byzantine failures}
\label{subsec:byzantine}

Byzantine failures are the most malicious failures that can occur in a distributed system. They were defined by Lamport in the famous \cite{byzantine} .

The adjective “byzantine” comes from an allegory to describe the problem, where several divisions of the byzantine army are camped outside an enemy city. Each division obeys its general and generals can communicate together via messenger. Their goal is to plan an attack. The notion of consensus comes from the need to plan something together.

However, not all generals are considered loyal. Some of them might be traitors to the cause. Indeed the generals must have a algorithm taking into account the possibility of traitor generals (“byzantine” processes in the system) to achieve the consensus (all loyal generals agree on the plan).

Formally, a byzantine process can have any arbitrary behaviour. We only assume that the receiver knows the identity of the sender of a message.

In the paper, the problem is decribed as so :
\subparagraph{Setting}
Given a system of n components, t of which are dishonest, and assuming only point-to-point channel between all the components.
Whenever a component A tries to broadcast a value x, the other components are allowed to discuss with each other and verify the consistency of A's broadcast, and eventually settle on a common value y.
\subparagraph{Property}
The system is said to resist Byzantine faults if a component A can broadcast a value x, and then:
If A is honest, then all honest components agree on the value x.
In any case, all honest components agree on the same value y.
\subparagraph{Variants}
The problem has been studied in the case of both synchronous and asynchronous communications.

The paper studies a few examples, and contains demonstrations for important results.  For example, no solution can be found for oral messages if there are 3 generals, one of whom is a traitor.

Byzantine failures are studied in systems with and without authentication. With authentication, messages can be signed with the name of the sending processor. The signature is presumed impossible to counterfeit. Without authentication, there is no guarantee regarding authentication for byzantine processes.

\subsection{Consensus protocols existence boundaries}

To conclude on the possibilities to reach consensus depending on our system of processes, the type of fault that we want to study, the kind of consensus that we want and the level of synchrony that we assume,
we display a table from \cite{partialSynchrony} with the smallest number of processors Nmin for which a t-resilient consensus protocol exists. ( t being the number of faulty processes)

\begin{center}
    \captionof{table}{SMALLEST NUMBER OF PROCESSORS $N_{min}$ FOR WHICH A t-resilient CONSENSUS PROTOCOL EXISTS}
    \begin{tabular}{ |m{2.4cm}|m{2.4cm}|m{2.4cm}|m{2.4cm}|m{2.4cm}|m{2.4cm}| }
        \hline
        Failure Type            & Synchronous & Asynchronous & Partially synchronous communication and synchronous processors & Partially synchronous communication and processors & Synchronous communication and partially synchronous processors \\
        \hline\hline
        Fail-Stop (crash)       & t           & \( \infty \) & 2t+1                                                           & 2t+1                                               & t                                                              \\
        \hline
        Omission                & t           & \( \infty \) & 2t+1                                                           & 2t+1                                               & [2t, 2t+1]                                                     \\
        \hline
        Authenticated Byzantine & t           & \( \infty \) & 3t+1                                                           & 3t+1                                               & 2t+1                                                           \\
        \hline
        Byzantine               & 3t+1        & \( \infty \) & 3t+1                                                           & 3t+1                                               & 3t+1                                                           \\
        \hline
    \end{tabular}
\end{center}

\(Nmin = \infty\) means that irrevocable consensus is a priori impossible with these hypothesis \cite{fischerlp}
The closed interval \([2t, 2t+1]\) means that  \(2t \leqslant Nmin \leqslant 2t+1\)

\subsection{Performance}

\paragraph{Performance evaluation of a consensus algorithm}

One way of evaluating consensus protocols is to assess, for one system of processes, whether or not they resist a fail-stop failure, an omission failure and/or a byzantine failure. But if we want to be more precise in our assessment, for example to compare multiple Byzantine fault-tolerant protocol we need more criteria.

Studying the performance of consensus algorithms is not easy, as reported in \cite{analysisperf} (“Consensus protocols are typically too complex for analytical approaches to performance evalution”) and \cite{failureDetection} limits the study to the pure speed of the protocol (providing that the consensus is indeed reached). According to Urban and Schiper \cite{comparing},

\begin{displayquote}
    Most of the time, consensus algorithms are evaluated using
    simple metrics like time complexity (number of communication steps) and message complexity (number of messages). This gives, however, little information on the real
    performance of those algorithms. […]

    Most papers on the performance of agreement algorithms only consider failure free executions, which only gives a partial and incomplete
    understanding of the behavior of the algorithms
\end{displayquote}
They suggest a benchmark based on measuring what they call the early latency of atomic broadcast.  This value is measured under multiple conditions of workload (how much the system has to process) and faultloads (how many failing processors the system has).

Some protocols, like Bitcoin, implement Proof-of-Work as a part of their consensus algorithm, and we will discuss why Proof-of-Work is used, but many publications (like \cite{decarbonizing} ) raised the issue of its electricity consumption. Some even track the electric consumption of the network on a day to day basis \cite{cambridgeIndex}. So energy consumption is a criteria to keep in mind when studying consensus algorithms.
There are basically two states : either the consensus algorithm implements Proof-of-Work, or not and if not then the electricity consumption will be like that of any server, depending on the number of nodes participating.

\paragraph{Using the state machine approach}

Even if some papers had described the concept years before, the state-machine approach is first defined by Schneider \cite{stateMachine}.

\begin{displayquote}
    The state machine approach is a general method for implementing a fault-tolerant service by replicating servers and coordinating client interactions with server replicas.
\end{displayquote}

Using the state machine approach allows us to better assess the performance of consensus algorithm, because they are put in a situation where testing is easier : “Do I get the expected result when there is a crash fault / a byzantine fault on 1/2/n nodes in my system using this algorithm ? How long does it take ? ”

The Andrew benchmark \cite{scaleperf} for example is used in the PBFT\footnote{Practical Byzantine Fault-Tolerance} paper \cite{pbft} to state that their replicated file system supporting NFS\footnote{Network File System} adds very little time cost compared to a standard NFS deamon.

\section{Review of some of the most popular consensus algorithms}
\subsection{Paxos}

Paxos was presented by Lamport \cite{parliament} to be a fault-tolerant consensus protocol. It can cope with processors experiencing failures, as long as they are not byzantine failures. Other assumptions are that the functioning processors are connected to each other. No speed requirement is done : processors may operate at any speed, and the same goes with connection in the network : messages are sent asynchronously and might take a long time to be delivered, or even get lost.

As we saw earlier in \cite{fischerlp}, in a completely asynchronous environment, progress is never guaranteed, even if only one processor fails. However, by measuring time it is easy to fall in the “partial synchrony” case of our table. The conditions that would prevent Paxos from making progress are rarely met in practice. In the paper, it is said that

\begin{displayquote}
    Achieving the progress condition required that legislators be able to measure the passage of time.
\end{displayquote}

Paxos is now considered a family of protocols because of the many variants that have been created over time. We will first review the most basic version of it and then some of the developments that can be made.

\paragraph{Vanilla Paxos}

In the Paxos algorithm, each processor can play a role at a certain time. Those roles are Acceptor (voters), Proposer, Learner and Leader

In the Part-Time Parliament, multiples protocols are described, always in the metaphoric description of the parliament. Priests ($p$) in the parliament are the processors in the protocol.
The first protocol is called the preliminary protocol but following this one a second protocol called the basic protocol is the one who stayed as the Paxos algorithm.
It is easier to implement than the preliminary protocol because  the priests (the processors) have to keep less information in their ledger (in memory ) to be a part of the protocol.

Those informations are (for a priest $p$) :
\begin{enumerate}
    \item $lastTried[p]$ : The number of the last ballot that $p$ tried to initiate, or $-\infty$ if there was none.
    \item $prevVote[p]$ : The vote cast by $p$ in the highest-numbered ballot in which is voted, or $-\infty$ if he never voted
    \item $nextBal[p]$ : The largest value of $b$ for which $p$ has sent a \(LastVote(b, v)\) message, or $-\infty$ if he has never sent such a message.
\end{enumerate}

The basic protocol ( which I won't copy here ) guarantees consistency, which for recall means no two ledgers can contain contradictory information (different decrees with the same number); but it doesn't guarantee progress because it only states what the priest can do, not what they have to do.
In order to achieve progress, Lamport introduces the Complete Synod Protocol, where priest must complete some of the steps of the protocol as soon as they can.
With a few hypothesis on priest reacting to messages and messages being delivered, the Complete Synod Protocol guarantees progress if and only if only one priest who doesn't live the chamber propose new decrees.
The complete Paxos protocol therefore includes a process to select a leader (the president). The president selection requirement is the following :

\begin{displayquote}
    If no one entered or left the Chamber, then after T minutes exactly one priest in the Chamber would consider himself to be the president.
\end{displayquote}

Some details are given on how the new president becomes the new president, but it is similar to a decree proposal process.

\paragraph{Variants}

Multi-Paxos is the main extent of Paxos, introduced in the Part-Time Parliament as the Multi-decree parliament.
It addresses passing multiple decrees instead of just focusing on passing one. This leads to some improvements in the runs, compared to running the basic Paxos time and time again.
For example, if the leader is elected in one Paxos run and does not fail, he can be used in the next run as well, without compromising any of the properties of Paxos.

Let's say we use Paxos to provide access to a distributed state machine.
Therefore, computing a Paxos run each time the service is queried by a client would result in a significant amount of unecessary processing, compared to a leader-based process with Paxos runs only to tackle failures.
Most of the time, variants of Paxos would be implemented, not the original Paxos mentionned as the Basic Protocol.

Some people have since worked on other variants. Lamport himself worked on a Fast Paxos algorithm \cite{fastpaxos} (basically reducing the number of message delays when the information comes from a client query).

\subsection{Raft}

The Raft algorithm was introduced in In search for an understandable consensus algorithm \cite{understandable}.
The Raft is presented as as efficient as Paxos and producing a result equivalent to (multi-)Paxos.
It is also supposed to be more understandable than Paxos, and with a different structure. Just like Paxos, the Raft can handle stop failures, but not byzantine failures.

Raft basically relies on a stronger leader than Paxos (in which the leader is introduced only for performance purpose) with followers only responding to the leader's request as long as there is no failure.
The third state (leader and follower being the two first) is candidate and is used in the case of a leader failure triggering a new leader election.

\subsection{PBFT}

PBFT, standing for Practical Byzantine Fault Tolerance is a consensus algorithm introduced in 1999 by Miguel Castro and Barbara Liskov \cite{pbft}

PBFT is described in the context of the state machine approach, which means that the goal is a bit more than just reach consensus between the nodes, but we will here isolate the consensus algorithm laying behind the description of the paper. The algorithm improves protocols presented in other papers, Rampart and SecureRing, which relied on the synchrony assumption.

The algorithm tolerates f faulty nodes if there are 3f+1 nodes in the system, and as said before adds very little time cost compared to a standard implementation of the same service (NFS in this case).

PFBT uses cryptography to authenticate the nodes in the exchange. In the basic version (section 4 of the paper), digital signatures are used to authenticate all the messages. In the final version described in the paper however, in order to optimize performance, digital signatures are used only for view-change and new-view messages and all other messages are authenticated using message authentication code (MACs).

\subsection{Proof-of-Work}
\label{sec:pow}

Even if it is often refered as so, Proof-of-Work is not in itself a consensus algorithm. Proof-of-Work was at inception a denial-of-service counter-measure \cite{hashcash}. Proof-of-Work algorithms allow users to show that they computed some amount of work, for example when using a service.

Without Proof-of-Work, an attacker can potentially send millions of request to a service that has to process them in what is known as a denial-of-service attack. If the services requieres the request to contains a Proof-of-Work stamp, the attacker would need for example one second to compute each stamp, which would need a lot of power and time to do the attack, in practice avoiding this kind of attacks completely.

In the bitcoin blockchain \cite{bitcoin}, Proof-of-Work is used to build the consensus algorithm which consists in giving trust to the longest chain\footnote{Longest chain of blocks in the blockchain, see \cite{blockchain} or \cite{bitcoin} for details}. This is some times referred to as the Nakamoto consensus, like in \cite{proconsensus}. The longest chain can only be built if you have the majority of computing power on the network. 
Economical rationality makes it that nodes want to play with the rules to win the newly created coins (when one block is mined, the successful node (miner) gets a compensation for his work), therefore “avoiding” the byzantine generals problem.

The bitcoin network is permissionless, and theoretically, any number of byzantine nodes (without computing power) can join the network without affecting its security.

I included Proof-of-Work because it is the technology that allowed all the developments of the crypto world but studying it deeply is beyond the scope of this project.

\subsection{Proof-of-Stake}
The first occurrence of Proof-of-Stake can be found behind the cryptocurrency PPCoin (or PeerCoin) \cite{ppcoin}, a project receiving very little interest nowadays. However, the technology gained popularity over time and nowadays much more cryptocurrencies are based on Proof-of-Stake than on Proof-of-Work \cite{coinmarketcap}. 
The second biggest cryptocurrency by market cap\footnote{Market capitalization (or market cap) is the total value of all the coins that have been mined. It is calculated by multiplying the number of coins in circulation by the current market price of a single coin.}, Ethereum, even moved from Proof-of-Work to Proof-of-Stake not long before these lignes are written, in on of the biggest event the world of crypto has ever seen (known as The Merge).

Proof-of-Stake consists in replacing the trust from computing power with the trust from holding the coins of the ecosystem, therefore prooving that someone has something to loose beeing dishonest.

Again, I'm not going to go further in the details of the different implementations of Proof-of-Stake, how inflation is managed in these ecosystems, that is a whole other topic. The Ethereum Foundation claims that Proof-of-Stake is more secure, less energy-intensive, and better for implementing new scaling solutions \cite{ethereum}. At least, it tackles the biggest criticism on cryptocurrencies from the mainstream media, which is that Proof-of-Work uses too much power \cite{bitcoinUse}.

\subsection{Avalanche}

Avalanche introduces a new family of Byzantine Fault Tolerant Protocols \cite{snowprotocol}. The Avalanche blockchain which relies on the Snowball protocol reached 30 billion USD market cap at its peak. Avalanche consistently stands in the top 10 public blockchains since 2020.

Slush, Snowflake, Snowball are the three main variants that are presented, each being more complex than the previous one.

Unlike Paxos, Raft and PBFT, in which all correct nodes reach the consensus with absolute certainty the Snow protocol family, like the Nakamoto consensus, may revert with some probability ε. It is not an irrevocable consensus algorithm.

I won't get here into the details of the algorithm and why it works. Basically each node, when it receives a new state, updates it own state with the information and picks a small sample of the network to propagate its new own state. The details vary between Slush (non BFT), Snowflake (BFT) and Snowball (adding “confidence”).

\section{Conclusion}

The question of reaching consensus among a set of distributed nodes in a vast topic, which started from distributed computing and distributed database mostly to prevent hardware failures and scale infrastructures. Nowadays most of the research is done around blockchains and how to prevent Byzantine failures with fast and secure consensus protocols.

Sacrificing the assurance of reaching a irrevocable consensus for a probabilistic approach seems to be the norm among public blockchains. Indeed, as they are public, open and permissionless, having one node trying to reach consensus by sending requests to a quorum of nodes which have to validate, like Paxos algorithms do is too slow and heavy.

\bibliographystyle{alpha}
\bibliography{biblio}

\end{document}

### 2022-10-03 biblio-rp.tex
typos

@@
 The second biggest cryptocurrency by market cap\footnote{Market capitalization (or market cap) is the total value of all the coins that have been mined.
-It is calculated by multiplying the number of coins in circulation by the current market price of a single coin.}, Ethereum, even moved from Proof-of-Work to Proof-of-Stake not long before these lignes are written, in on of the biggest event the world of crypto has ever seen (known as The Merge).
+It is calculated by multiplying the number of coins in circulation by the current market price of a single coin.}, Ethereum, even moved from Proof-of-Work to Proof-of-Stake not long before these lignes were written, in one of the biggest event the world of crypto has ever seen (known as The Merge).
 
-Proof-of-Stake consists in replacing the trust from computing power with the trust from holding the coins of the ecosystem, therefore prooving that someone has something to loose beeing dishonest.
+Proof-of-Stake consists in replacing the trust from computing power with the trust from holding the coins of the ecosystem, therefore prooving that someone has something to loose by beeing dishonest.
 

### 2022-10-29 scripts/markovBeta.py
parrallelized code to computed probabilities

@@ -3,140 +3,177 @@
 import math
 import numpy as np
 import sys
+from joblib import Parallel, delayed
 
 # constants
-k = 12  # how many requests per group of request
-n = 20   # number of nodes
+# k = 8  # how many requests per group of request
+# n = 30   # number of nodes
 
 # Probabilities : array[i][j][k][l] probability of i colored nodes, j semi-colored nodes, k requests from colored nodes, l requests from semi-colored nodes
 
-array = np.ndarray(shape=(n+1, n+1, n+1, n+1), dtype=float)
-array.fill(-1)  # table contains -1 if the probability was not yet computed
+# array = np.ndarray(shape=(n+1, n+1, n+1, n+1), dtype=float)
+# array.fill(-1)  # table contains -1 if the probability was not yet computed
 
-# one node receives the information and emits one group of requests
-array[0][1][0][1] = 1
-print(array)
-
-
-def Pcreqs(c, s, reqsc, reqss):
-    print("requested proba of " + str(c) + " colored nodes, " + str(s) + " semi colored nodes, " + str(reqsc) + " groups of requests from colored nodes and " +
-          str(reqss) + " groups of requests from semi-colored nodes")
-    if c + s < 1:
-        return 0
-    if c < 0:
-        return 0
-    if c+s > n:
-        return 0
-    if s < 0:
-        return 0
-    if reqsc > c:
-        return 0
-    if reqsc < 0:
-        return 0
-    if reqss > s+c:
-        return 0
-    if reqss < 0:
-        return 0
-
-    if array[c][s][reqsc][reqss] != -1:
-        print(
-            "P(" + " ".join([str(c), str(s), str(reqsc), str(reqss)]) + ") = " + str(array[c][s][reqsc][reqss]))
-        return array[c][s][reqsc][reqss]
-
-    # considering a step with a group of requests from a semi-colored node
-    print("considering a step with a group from sm node")
-    ress = 0
-    for tv in range(k+1):
-        for ts in range(k+1):
-            if n-c-s+ts+tv < tv:
-                continue
-            if s-tv-1+ts < ts:
-                continue
-            if k-tv-ts > c-ts:
-                continue
-            if c-ts < 0:
-                continue
-            if s-tv-1 < 0:
-                continue
-            if n-c-s+ts+tv < 0:
-                continue
-            if k - tv-ts < 0:
-                continue
-            if reqss+1-tv < 0:
-                continue
-            if reqsc-ts < 0:
[…]

### 2022-10-29 scripts/markovBeta.py
display 3d graph of epsilon depending on k and n

@@ -4,6 +4,8 @@ import math
 import numpy as np
 import sys
 from joblib import Parallel, delayed
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
 
 # constants
 # k = 8  # how many requests per group of request
@@ -170,10 +172,20 @@ def computeAll(n, ks):
         if n > k:
             resn.append(getNoConsensusProb(n, k))
         else:
-            resn.append(-1)
+            resn.append(0)
     return resn
 
 
 results = Parallel(n_jobs=-1)(delayed(computeAll)(n, ks) for n in nodes)
 
-print(results)
+X, Y = np.meshgrid(ks, nodes)
+
+fig = plt.figure()
+
+ax = plt.axes(projection="3d")
+ax.contour3D(X, Y, results, 80, cmap="viridis")
+ax.set_xlabel("k")
+ax.set_ylabel("n")
+ax.set_zlabel("epsilon")
+
+plt.show()

### 2022-10-31 scripts/markovBeta.py
more pitch, log epsilon in graph

@@ -170,9 +170,13 @@ def computeAll(n, ks):
     resn = []
     for k in ks:
         if n > k:
-            resn.append(getNoConsensusProb(n, k))
+            g = getNoConsensusProb(n, k)
+            if g == 0:
+                resn.append(100)
+            else:
+                resn.append(-np.log(getNoConsensusProb(n, k)))
         else:
-            resn.append(0)
+            resn.append(100)
     return resn
 
 
@@ -183,9 +187,9 @@ X, Y = np.meshgrid(ks, nodes)
 fig = plt.figure()
 
 ax = plt.axes(projection="3d")
-ax.contour3D(X, Y, results, 80, cmap="viridis")
+ax.contour3D(X, Y, results, 200, cmap="viridis")
 ax.set_xlabel("k")
 ax.set_ylabel("n")
-ax.set_zlabel("epsilon")
-
+ax.set_zlabel("-log(epsilon)")
+ax.set_title("Probability of not reaching consensus")
 plt.show()

### 2022-10-31 article/article.tex
article framework

\documentclass[11pt, twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{fdsymbol}
\usepackage{textgreek}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{url}
\usepackage{array}
\usepackage{csquotes}
\usepackage{caption}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}

\title{Decentralized state machine using Nakamoto (probabilistic) consensus\\\medskip Research Project 2022}
\author{Nicolas COURAUD (nicolas.couraud@etu.emse.fr)\\École Nationale Supérieure des Mines de Saint-Étienne}

\begin{document}

\maketitle
\onecolumn
\section*{Abstract}

As defined by Schneider \cite{stateMachine}, the state machine approach is a general method for implementing a fault-tolerant service
by replicating servers and coordinating client interactions with server replicas.

In this article, we look at state machines as a way to hold in multiple servers copies of the same record. That record, in our implementation, is a standard 
key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).

We show how one can build a decentralized state machine using Nakamoto Consensus (in our implementation we used the Snowball consensus algorithm), and we study the advantages 
and drawbacks of such an approach, compared to classical consensus algorithms like Raft \cite{understandable}.

The implementation of the project is available at \href{https://github.com/Nicolascrd/distributed-state-machine}{github.com/Nicolascrd/distributed-state-machine}

\tableofcontents
\section{Traditional state machine replication}

\subsection{Introduction}

Introduction
Introduction
Introduction
Introduction
Introduction
Introduction
Introduction

\subsection{Byzantine fault-tolerance with classical consensus}

Byzantine
Byzantine
Byzantine
Byzantine
Byzantine
Byzantine
Byzantine
Byzantine

\section{Nakamoto (probabilistic) state machine replication}
\subsection{The Nakamoto Consensus}

The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus
The Naka consensus

\subsection{My Nakamoto state machine implementation}

Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine
Nakamoto state machine

\section{Performance Evaluation}
\subsection{Probability of not reaching consensus in Nakamoto Consensus}

Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov
Markov

\subsection{Number of requests required in regular functioning}

Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests
Number of requests

\subsection{In case of a failure}

Failure ?
Failure ?
Failure ?
Failure ?
Failure ?
Failure ?
Failure ?
Failure ?
Failure ?
Failure ?

\section{Conclusion}

Conclusion
Conclusion
Conclusion
Conclusion
Conclusion
Conclusion
Conclusion
Conclusion
Conclusion

\bibliographystyle{alpha}
\bibliography{biblio}

\end{document}

### 2022-10-31 article/article.tex
part 1 article

@@
 
-In this article, we look at state machines as a way to hold in multiple servers copies of the same record. That record, in our implementation, is a standard 
+In this article, we look at state machines as a way to hold in multiple servers copies of the same record. That record, in my implementation, is a standard 
 key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).
 
-We show how one can build a decentralized state machine using Nakamoto Consensus (in our implementation we used the Snowball consensus algorithm), and we study the advantages 
+I show how one can build a decentralized state machine using Nakamoto Consensus (in our implementation we used the Snowball consensus algorithm), and I study the advantages 
 and drawbacks of such an approach, compared to classical consensus algorithms like Raft \cite{understandable}.
@@
 
-Introduction
-Introduction
-Introduction
-Introduction
-Introduction
-Introduction
-Introduction
+Decentralized state machine systems are currently built, for the most part, around classical consensus algorithms.
 
-\subsection{Byzantine fault-tolerance with classical consensus}
+In classical consensus, all of the nodes have to know each other. It is a permissioned model, which means that nodes have to get an approval to get into the network. 
+That is very normal for private infrastructure but might not work for public infrastructure. Because all nodes must communicate with each other, the communication cost in generally considered high.
 
-Byzantine
-Byzantine
-Byzantine
-Byzantine
-Byzantine
-Byzantine
-Byzantine
-Byzantine
+Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}.
+In my project, I used the Raft consensus algorithm \cite{understandable} because it has the same performances
+as Paxos, and is simpler to understand and implement.
+
+In Raft, each node can be either Leader, Follower or Candidate. In regular functioning, one node will be the leader and all the others will follow. 
+The Leader node sends regular heartbeat requests to notify followers that it is still running. If they stop receiving the heartbeat, they switch to the follower status.
+
+The client can ask any node to add a log to the record. 
+The request includes the number of the log that the client wants to add and the log himself.
+If the client as a node which is not the leader, the request is transfered to the leader.
+If there is already a record at that number,
+the leader returns an error message to the client. Otherwise, the leader updates its record and asks all his followers to update themselves as well.
+
+When requesting a log, the client can also query any node in the network. The node just responds with the log at that number in the local record that it is holding.
+
+The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean.
+If True, if the leader sends heartbeats and a node does not reply, it will be eliminated from the network in the knowledge of all the nodes.
+This make it possible to
+crash a majority of nodes and still have the state machine running and usable on all the surviving nodes.
+
+\subsection{Byzantine fault-tolerance (BFT) with classical consensus}
+
+Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way.
+In practice, it means that the node will stay up and can send malicious requests to
+interfere with the functioning of the network.
+
+My implementation of the state machine does not tolerate Byzantine Faults at all.
+With only one byzantine node, it already endangers the whole system.
+Indeed, the malicious node can "elect himself" in Raft, and then send heartbeats to all
+the node in the network, including the legitimate leader to become the leader de facto. Then, it can ignore requests, reply whatever it wants and write any log it wants in any node it wants.
+
+To tackle this issue, BFT algorithms were designed \cite{pbft}.
+They include a layer of authentification, but are not necessarily much slower.
+However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
+For example, assuming partial synchrony of the nodes, consensus protocols can handle t crash failures with 2t+1 nodes but t byzantine failures with no less than 3t+1 nodes.
 

### 2022-11-01 article/article.tex
more article

@@
 \usepackage{amsmath, amssymb}
+\usepackage{mathtools}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+\usepackage{nameref}
 \usepackage{url}
@@
 \usepackage{caption}
+\usepackage{algpseudocode}
+\usepackage[toc,page]{appendix}
+\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
+\providecommand{\myceil}[1]{\left \lceil #1 \right \rceil }
 \hypersetup{
@@
 \tableofcontents
+
 \section{Traditional state machine replication}
@@
 
-To tackle this issue, BFT algorithms were designed \cite{pbft}.
+To tackle this issue, BFT algorithms were designed (such as \cite{pbft}).
 They include a layer of authentification, but are not necessarily much slower.
@@
 
-The Naka consensus
-The Naka consensus
-The Naka consensus
-The Naka consensus
-The Naka consensus
-The Naka consensus
-The Naka consensus
+Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
+Of course, ε can be so low that crucial systems can be built relying Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
+Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.
+
+Nakamoto consensus definetly shaked the consensus space in computer science with cryptocurrencies. In a different context, can it be useful as well ?
 
@@
 
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
-Nakamoto state machine
+To implement Nakamoto consensus, I chose the Snowball consensus algorithm \cite{snowprotocol}, which underlies the Avalanche blockchain.
+
+The goal of snowball is to reach binary consensus starting from a network of nodes.
+
+I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (because it is the basis of Avalanche, which is that payment system).
+Snowball is built by iterations, but I will only focus on Snowball, the last iteration, because it is the most interesting and the one I tested the performances in the end.
+Slush and Snowflake are interesting to understand
+where we are coming from, but no more.
+
+I adapted Snowball in the following ways : Each node, in addition to having all the snowball consensus related data (not much), has a version of the record. All nodes starts with a empty record.
+Because it is the requirement for a decentralized state machine, each node can be queried to add a log to the record. Each node can also be queried to get one log in the record.
+
+The snowball protocol is triggered when the client asks to add a new log to an index that has not been touched yet.
+Indeed, the system works as if each index in the record corresponds to an independant snowball run.
+
+In snowball, each node has a counter of "confidence" which can tip towards one way or another.
+In my implementation, there is no binary consensus because the only value the system has is the log request.
+When receiving an add-log request, the node follows this sequence :
+\begin{algorithmic}
+    \State from request (client or other node):
+    \State $logToAdd$ (string)
+    \State $logPosition$ (int)
+
+    \State \\from node memory :
+    \State $record$ (map[int]string)
+    \State $cnt$ (map[int]int)
+    \\
+    \If{$record[logPosition]$ exists}
+    \State reply $record[logPosition] == logToAdd$
+    \Else
+    \State $record[logPosition] \gets logToAdd$
+    \State reply $True$
+    \State initiateRequest($logToAdd$, $logPosition$)
+    \EndIf
+
+\end{algorithmic}
+
+initiateRequest corresponds to the following sequence :
+
+\begin{algorithmic}
+    \While{$True$}
+    \State $success \gets initiateQuery(req)$
+    \State $initiateQuery$ picks a sample in the network and queries it
+    \If{$success$}
+    \State $cnt[logPosition]++$
+    \Else
+    \If{$cnt[logPosition] == 1$}
+    \State delete($record$, $logPosition$)
+    \State $cnt[logPosition] \gets 0$
+    \Else
+    \State $cnt[logPosition]--$
+    \EndIf
+    \EndIf
+    \If{$cnt[logPosition] \geqslant CounterThreshold$}
+    \State $record[logPosition] \gets logToAdd$
+    \State return
+    \EndIf
+    \EndWhile
+\end{algorithmic}
+
+3 parameters are used in my implementation :
+\begin{itemize}
+    \item sampleSize (k) : the number of nodes that are queried each round
+    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
+          (with \begin{equation*} \alpha \geq \floor{k/2}  \end{equation*})
+    \item counterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
+\end{itemize}
 
@@
 
-Markov
-Markov
-Markov
-Markov
-Markov
-Markov
-Markov
-Markov
-Markov
-Markov
+In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set 
+of values (n, k, β). (α is not included because we assume no byzantine nodes in this part).
+
+The evolution of the system of nodes, for one position (one snowball run) is a Markov process. To compute to probability of not reaching consensus, we need to identify the possible states in the 
+Markov process, the possible transitions and the transitions probabilities.
+
+\paragraph{β = 1}
+With β = 1, the state is (c, reqs) with :
+\begin{itemize}
+    \item c = Number of colored nodes ($c \leq n$)
+    \item reqs = Number of groups of requests which are queried for next round. (They all come from neutral nodes)
+\end{itemize}
+If the state converges to $c = n$, consensus is reached.
+If the state converges to $reqs = 0$, with $c \neq 0$, consensus is not reached when the snowball run is done.
+
+Then, the probability of reaching a certain state (1) is the sum of the probability of a different state (2) times the probability of going from state 2 to state 1.
+That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.
+
+Therefore, the probability of being at state (c, reqs) is (the term in the sum is considered 0 if it does not make sense, which is the case for some t) :
+The sum represents a step of the resolution of one group of requests (t represent the number of neutral nodes touched)
+\begin{equation*}
+    P(c, reqs) = \sum_{t=0}^{k+1} {n-c+t \choose t}{c-t-1 \choose k-t}P(c-t, reqs+1-t)
+\end{equation*}
+
+\paragraph{β = 2}
+With β = 2, the state is (c, s, rc, rs), with:
+\begin{itemize}
+    \item c = Number of colored nodes (cnt = 2)
+    \item s = Number of semi-colored nodes (cnt = 1)
+    \item rc = Number of groups of requests from colored nodes
+    \item rs = Number of groups of requests from semi-colored nodes
+\end{itemize}
+
+We must separate requests in those two groups because a request cannot be directed to the node initiating it. 
+Therefore the probabilities of hitting a colored node is smaller if the request comes from a colored node than from a semi-colored node.
+
+The term in the sums is considered 0 if it does not make sense.
+First sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
+Second sum represents a step of the resolution of one group of requests, coming from a colored node. 
+To do the calculation, we assume that requests from semi-colored nodes resolve faster.
+\begin{multline*}
+    P(c, s, rc, rs) = \\ \sum_{tv=0}^{k+1}\sum_{ts=0}^{k+1} {n-c-s+tv \choose tv}{s-tv-1+ts \choose ts}{c-ts \choose k-tv-ts}P(c-ts, s-tv+ts, rc-ts, rs+1-tv)
+    \\ + \sum_{ts=0}^{k+1} {n-c-s+rs \choose tv}{s-rs+ts \choose ts}{c-ts-1 \choose k-rs-ts}P(c-ts, s-rs+ts, rc+1-ts, 0)
+\end{multline*}
+\\
+With these formulas, we can compute the probability of not reaching consensus : view Appendices
 
@@
 
+\onecolumn
 \bibliographystyle{alpha}
@@
 
+\begin{appendices}
+    \label{Appendices}
+    \begin{figure*}
+        Probability of not reaching consensus (β = 1) : ε 
+        \centering
+        \includegraphics[width=16.5cm]{images/smallNoConsensusBetaOne.png}
+    \end{figure*}
+    \begin{figure*}
+        Probability of not reaching consensus (β = 1) : -log(ε) 
+        \centering
+        \includegraphics[width=16.5cm]{images/smalllogNoConsensusBetaOne.png}
+    \end{figure*}
+    \begin{figure*}
+        Probability of not reaching consensus (β = 2) : ε 
+        \centering
+        \includegraphics[width=16.5cm]{images/smallnoconsensus.png}
+    \end{figure*}
+    \begin{figure*}
+        Probability of not reaching consensus (β = 2) : -log(ε) 
+        \centering
+        \includegraphics[width=16.5cm]{images/smalllognoconsensus.png}
+    \end{figure*}
+\end{appendices}
 \end{document}

### 2022-11-01 scripts/markov.py

@@ -1,63 +1,82 @@
 import math
 import numpy as np
-
-# constants
-k = 12  # how many requests per group of request
-n = 20  # number of nodes
-
-# Probabilities : array[i][j] probability of i colored states, j groups of requests
-array = np.ndarray(shape=(n+1, n+1), dtype=float)
-array.fill(-1)  # table contains -1 if the probability was not yet computed
-
-for i in range(n+1):
-    array[0][i] = 0
-
-array[1][1] = 1  # one node receives the information and emits one group of requests
-print(array)
+from joblib import Parallel, delayed
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
 
 
-def Pcreqs(c, reqs):
-    print("requested proba of " + str(c) + " colored nodes and " +
-          str(reqs) + " groups of requests")
-    res = 0
-    if c < 1:
-        return 0
-    if c > n:
-        return 0
-    if reqs > c:
-        return 0
-    if reqs < 0:
-        return 0
+def getNoConsensusProb(n, k):
 
-    if array[c][reqs] != -1:
-        return array[c][reqs]
-    for t in range(k+1):
-        if 0 > k-t or k-t > c-t-1:
-            continue
-        if 0 > t or 0 > n-c:
-            continue
-        if 0 > k or k > n-1:
-            continue
+    array = np.ndarray(shape=(n+1, n+1), dtype=float)
+    array.fill(-1)  # table contains -1 if the probability was not yet computed
+    for i in range(n+1):
+        array[0][i] = 0
 
-        if reqs + 1 - t > 0: # no requests can't progress
-            res += math.comb(n-c+t, t) * math.comb(c-t-1, k-t) * \
-                Pcreqs(c-t, reqs+1-t)
-    ans = res / math.comb(n-1, k)
-    array[c][reqs] = ans
-    return ans
+    # one node receives the information and emits one group of requests
+    array[1][1] = 1
+
+    def Pcreqs(c, reqs):
+        res = 0
+        if c < 1:
+            return 0
+        if c > n:
+            return 0
+        if reqs > c:
+            return 0
+        if reqs < 0:
+            return 0
+
+        if array[c][reqs] != -1:
+            return array[c][reqs]
+        for t in range(k+1):
+            if 0 > k-t or k-t > c-t-1:
+                continue
[…]

### 2022-11-01 scripts/markovBeta.py

@@ -7,41 +7,7 @@ from joblib import Parallel, delayed
 import matplotlib.pyplot as plt
 from mpl_toolkits.mplot3d import Axes3D
 
-# constants
-# k = 8  # how many requests per group of request
-# n = 30   # number of nodes
 
-# Probabilities : array[i][j][k][l] probability of i colored nodes, j semi-colored nodes, k requests from colored nodes, l requests from semi-colored nodes
-
-# array = np.ndarray(shape=(n+1, n+1, n+1, n+1), dtype=float)
-# array.fill(-1)  # table contains -1 if the probability was not yet computed
-
-# # one node receives the information and emits one group of requests
-# array[0][1][0][1] = 1
-# print(array)
-
-
-# for i in range(n+1):
-#     for j in range(n+1):
-#         for u in range(n+1):
-#             for v in range(n+1):
-#                 array[i][j][u][v] = Pcreqs(i, j, u, v)
-
-# print(Pcreqs(0, 4, 0, 3))
-# print(Pcreqs(3, 1, 3, 2))
-# print(array)
-
-# sumNoRequests = 0
-# noConsensus = 0
-
-# for i in range(0, n+1):
-#     for j in range(0, n+1):
-#         sumNoRequests += array[i][j][0][0]
-#         if i+j != n:
-#             noConsensus += array[i][j][0][0]
-
-# print("SUM", sumNoRequests)
-# print("no consensus", noConsensus)
 
 
 def getNoConsensusProb(n, k):
@@ -111,34 +77,32 @@ def getNoConsensusProb(n, k):
         # considering a step with a group of requests from a colored node
         # print("considering a step with a group from colored node only if no groups of requests from semi colored nodes")
         resc = 0
-        for tv in range(k+1):
-            if reqss != tv:
-                # no requests from semi colored nodes
+        tv = reqss
+        # no requests from semi colored nodes
+        for ts in range(k+1):
+            if n-c-s+ts+tv < tv:
                 continue
-            for ts in range(k+1):
-                if n-c-s+ts+tv < tv:
-                    continue
-                if s-tv+ts < ts:
-                    continue
-                if k-tv-ts > c-ts-1:
-                    continue
-                if c-ts-1 < 0:
-                    continue
-                if s-tv < 0:
-                    continue
-                if n-c-s+ts+tv < 0:
-                    continue
-                if k - tv-ts < 0:
-                    continue
-                if reqss-tv < 0:
[…]

### 2022-11-01 article/article.tex
more article

@@
 \usepackage{algpseudocode}
+\usepackage[a4paper, total={6.5in, 9in}]{geometry}
 \usepackage[toc,page]{appendix}
@@
 In my implementation, there is no binary consensus because the only value the system has is the log request.
-When receiving an add-log request, the node follows this sequence :
+When receiving an add-log request, the node follows this sequence, in which $logToAdd$($string$) and $logPosition$($int$) are variables from the request (client or other node) and $record$($map[int]string$) and $cnt$($map[int]int$) are variables from the node memory:
 \begin{algorithmic}
-    \State from request (client or other node):
-    \State $logToAdd$ (string)
-    \State $logPosition$ (int)
-
-    \State \\from node memory :
-    \State $record$ (map[int]string)
-    \State $cnt$ (map[int]int)
-    \\
     \If{$record[logPosition]$ exists}
@@
     \EndIf
-
 \end{algorithmic}
@@
     \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
-          (with \begin{equation*} \alpha \geq \floor{k/2}  \end{equation*})
+          (with $\alpha \geq \lceil k/2 \rceil$)
     \item counterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
@@
 \begin{itemize}
-    \item c = Number of colored nodes ($c \leq n$)
+    \item c = Number of colored (cnt = 1) nodes ($c \leq n$)
     \item reqs = Number of groups of requests which are queried for next round. (They all come from neutral nodes)
@@
 Therefore, the probability of being at state (c, reqs) is (the term in the sum is considered 0 if it does not make sense, which is the case for some t) :
-The sum represents a step of the resolution of one group of requests (t represent the number of neutral nodes touched)
 \begin{equation*}
@@
 \end{equation*}
+The sum represents the resolution of one group of requests of the consensus protocol for one position (t represent the number of neutral nodes touched)
 
@@
 \begin{itemize}
-    \item c = Number of colored nodes (cnt = 2)
-    \item s = Number of semi-colored nodes (cnt = 1)
+    \item c = Number of colored (cnt = 2) nodes
+    \item s = Number of semi-colored (cnt = 1) nodes
     \item rc = Number of groups of requests from colored nodes
@@
 With these formulas, we can compute the probability of not reaching consensus : view Appendices
+The scripts which were used to compute the graphs in Appendix are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}
 
@@
 
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
-Number of requests
+\paragraph{Raft (classical)}
+When the leader is up and giving instruction to all the followers, followers don't have to initiate requests.
+Therefore in normal functioning, the leader node will make $n-1$ requests every heartbeat.
+In addition to this, every client request will trigger $n-1$ requests from the leader, to update followers. If the request was directed from the client to a follower, one request is initiated
+by the follower to transfer the request to the leader.
+
+\paragraph{Snowball (Nakamoto)}
+The number of requests required to reach consensus on an add-log request is $k*\beta$ for each node.
 
 \subsection{In case of a failure}
+\paragraph{Fail-stop in Raft}
+If a follower crashes, the leader stays in place.
+If updateSystem is activated, one request will be made to each follower to update its knowledge of the system.
+That represents $n-2$ requests in this case
 
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
-Failure ?
+If the leader crashes, nodes will switch from the follower status to the candidate status, because they don't receive the heartbeat.
+However, the number of requests that will be made to elect a new leader depends on a number of factors.
+With a few nodes, the process happens with only one request : the first node to tick (notice that the heartbeat was not received) switches to candidate status, triggers a round of vote, collects all the votes and establishes itself as the new leader.
+In the meantime, the other nodes did not switch to candidate and start receiving the heartbeats from the new leader.
+With many nodes, nodes might tick "at the same time" and therefore candidate at the same time which can trigger many requests.
+To avoid an infinite escalation of currentTerm, which is the value incremented by each candidate, the ticker is supposed to happen randomly in some time window.
+
+Overall the Raft is very efficient at managing crash failures.
+
+\paragraph{Fail-stop in Snowball}
+If a node crashs in Snowball, some add-log request will not find a majority and therefore the consensus will be slower and require more requests.
+However, in production, the Snowball state-machine would include a mechanism to update the network knowledge among members, which can take the form of a Snowball run, and therefore continue as if no node had failed.
+Snowball is also very efficient at managing crash failures.
+
+\paragraph{Byzantine failures}
+The architecture required to test properly my implementations for byzantine failures goes beyong this project. But we can still think about the resilience of each implementation.
+With Raft, a byzantine node can easily take the control on the network by becoming the leader.
+It only needs to start as a follower, choose a currentTerm bigger than the current one and ask for votes, which other nodes will give.
+Once elected leader, the byzantine node can implement any information among followers. The state-machine is unusable and even potentially dangerously infected.
+
+With Avalanche, to begin with, a node cannot easily change the information that is already in place among nodes.
+This would require the consensus to switch from one way to the other, which is nearly impossible.
+However, in my implementation, the client cannot either change a log at one position.
+One thing the byzantine node cannot easily do in the Snowball implementation is register different informations among multiple nodes.
+Indeed, these nodes will want to verify the information by quering other nodes to reach consensus.
+The system will therefore go one way or the other.
 
@@
 
-Conclusion
-Conclusion
-Conclusion
-Conclusion
-Conclusion
-Conclusion
-Conclusion
-Conclusion
-Conclusion
+We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties.
+However, more research is required to use this approach at a greater scale.
+In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 
+Also, a comparison focused on Byzantine Fault Tolerance would be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
 
 \onecolumn
+
+\begin{appendices}
     \bibliographystyle{alpha}
     \bibliography{biblio}
-
-\begin{appendices}
     \label{Appendices}

### 2022-11-01 article/article.tex
typos

@@
 
-In this article, we look at state machines as a way to hold in multiple servers copies of the same record. That record, in my implementation, is a standard
+In this article, we look at state machines as a way to hold in multiple servers copies of the same record. In my implementation, this record is a standard
 key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).
@@
 Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}.
-In my project, I used the Raft consensus algorithm \cite{understandable} because it has the same performances
-as Paxos, and is simpler to understand and implement.
+In my project, I used the Raft consensus algorithm \cite{understandable} because it performs as well as Paxos, and is simpler to understand and implement.
 
@@
 
-The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean.
-If True, if the leader sends heartbeats and a node does not reply, it will be eliminated from the network in the knowledge of all the nodes.
-This make it possible to
+The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean. If $True$, if the leader sends heartbeats and a node does not reply, 
+it will be eliminated from the network in the knowledge of all the nodes. This makes it possible to
 crash a majority of nodes and still have the state machine running and usable on all the surviving nodes.
@@
 In practice, it means that the node will stay up and can send malicious requests to
-interfere with the functioning of the network.
+interfere with network usage.
 
-My implementation of the state machine does not tolerate Byzantine Faults at all.
-With only one byzantine node, it already endangers the whole system.
-Indeed, the malicious node can "elect himself" in Raft, and then send heartbeats to all
-the node in the network, including the legitimate leader to become the leader de facto. Then, it can ignore requests, reply whatever it wants and write any log it wants in any node it wants.
-
-To tackle this issue, BFT algorithms were designed (such as \cite{pbft}).
-They include a layer of authentification, but are not necessarily much slower.
+The Raft consensus algorithm is not BFT.
+BFT algorithms (such as \cite{pbft}) include a layer of authentification, but are not necessarily much slower.
 However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
-For example, assuming partial synchrony of the nodes, consensus protocols can handle t crash failures with 2t+1 nodes but t byzantine failures with no less than 3t+1 nodes.
+For example, assuming partial synchrony of the nodes, consensus protocols can handle $t$ crash failures with $2t+1$ nodes but $t$ byzantine failures with no less than $3t+1$ nodes.
 
@@
 Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
-Of course, ε can be so low that crucial systems can be built relying Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
+Of course, ε can be so low that crucial systems can be built relying on Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
 Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.
 
-Nakamoto consensus definetly shaked the consensus space in computer science with cryptocurrencies. In a different context, can it be useful as well ?
+Nakamoto consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state-machine with it ?
 
@@
 
-I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (because it is the basis of Avalanche, which is that payment system).
-Snowball is built by iterations, but I will only focus on Snowball, the last iteration, because it is the most interesting and the one I tested the performances in the end.
-Slush and Snowflake are interesting to understand
-where we are coming from, but no more.
+I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (it is the basis of Avalanche, which is that payment system).
+Snowball is built by iterations, but I will only focus on the last iteration, because it is the most interesting and the one I tested the performances in the end. 
+The other two algorithms, Slush and Snowflake are only interesting to understand the protocol.
 
@@
 
-In snowball, each node has a counter of "confidence" which can tip towards one way or another.
+In snowball, each node has a counter of "confidence" which can tip towards one way or the other.
 In my implementation, there is no binary consensus because the only value the system has is the log request.
-When receiving an add-log request, the node follows this sequence, in which $logToAdd$($string$) and $logPosition$($int$) are variables from the request (client or other node) and $record$($map[int]string$) and $cnt$($map[int]int$) are variables from the node memory:
+When receiving an add-log request, the node follows this sequence, in which $logToAdd$(string) and $logPosition$(int) are variables from the request (client or other node) and $record$(map[int]string) and $cnt$(map[int]int) are variables from the node memory:
 \begin{algorithmic}
@@
     \State $success \gets initiateQuery(req)$
-    \State $initiateQuery$ picks a sample in the network and queries it
+    \State ($initiateQuery$ picks a sample in the network and queries it)
     \If{$success$}
@@
           (with $\alpha \geq \lceil k/2 \rceil$)
-    \item counterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
+    \item CounterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
 \end{itemize}
@@
 In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
-of values (n, k, β). (α is not included because we assume no byzantine nodes in this part).
+of values (n, k, β). (α is not included because we assume no byzantine nodes in this section).
 
@@
 
-Then, the probability of reaching a certain state (1) is the sum of the probability of a different state (2) times the probability of going from state 2 to state 1.
+Then, the probability of reaching a certain state ($S1$) is the sum of the probability of a different state ($S2$) times the probability of going from state $S2$ to state $S1$.
 That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.
@@
 \\
-With these formulas, we can compute the probability of not reaching consensus : view Appendices
+With these formulas, we can compute the probability of not reaching consensus : view Appendices.
 The scripts which were used to compute the graphs in Appendix are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}
@@
 We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties. 
-However, more research is required to use this approach at a greater scale.
+In normal running state, the number of requests to reach consensus is a function of $k * \beta$ whereas in Raft, the leader must handle $n-1$ request on its own. 
+Raft has one consensus for the whole system, whereas in Snowball to each value correspond a different consensus. For most usage, classical consensus is probably the way to go.
+It would be useful to lead some more research on the topic. 
 In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 
-Also, a comparison focused on Byzantine Fault Tolerance would be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
+Also, a comparison focused on Byzantine Fault Tolerance would also be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
 

### 2022-11-02 article/article.tex
improving article

@@
 
-In this article, we look at state machines as a way to hold in multiple servers copies of the same record. In my implementation, this record is a standard
-key-value data structure (\href{https://go.dev/blog/maps}{a Go map}).
+In this article, we will look at a new family of probabilitic consensus algorithms, in the context of building a decentralized state machine.
 
-I show how one can build a decentralized state machine using Nakamoto Consensus (in our implementation we used the Snowball consensus algorithm), and I study the advantages
-and drawbacks of such an approach, compared to classical consensus algorithms like Raft \cite{understandable}.
+We will show that it is possible to build a decentralized state machine using Nakamoto Consensus, with the Snowball protocol \cite{snowprotocol} and will compare this approach to the traditional approach with
+a classical consensus algorithm (Raft \cite{understandable})
 
@@
 
-Decentralized state machine systems are currently built, for the most part, around classical consensus algorithms.
+Decentralized state machine systems are currently built around classical consensus algorithms.
 
+Classical consensus algorithms were designed in the late 20th century to tackle faulty hardware and network. Their goal is to avoid one service being made unavailable by a crash or a network failure. 
 In classical consensus, all of the nodes have to know each other. It is a permissioned model, which means that nodes have to get an approval to get into the network.
-That is very normal for private infrastructure but might not work for public infrastructure. Because all nodes must communicate with each other, the communication cost in generally considered high.
+Because all nodes must communicate with each other, the communication cost to reach consensus in generally considered high.
+That's why most of the time, the system is built with a leader/followers structure (it is the case with Raft).
+This makes the system efficient as one server can spread requests among followers to increase bandwidth and rely on them in the case of a crash.
+Also, while the leader is up it acts as a trusted party for both followers and the client, so it reduces the need to reach consensus at all time.
 
 Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}. 
-In my project, I used the Raft consensus algorithm \cite{understandable} because it performs as well as Paxos, and is simpler to understand and implement.
+\subsection{The Raft State Machine}
+In this project, I used the Raft consensus algorithm \cite{understandable} because it performs as well as Paxos, and is simpler to understand and implement.
 
 In Raft, each node can be either Leader, Follower or Candidate. In regular functioning, one node will be the leader and all the others will follow.
-The Leader node sends regular heartbeat requests to notify followers that it is still running. If they stop receiving the heartbeat, they switch to the follower status.
+The Leader node sends regular heartbeat requests to notify Followers that it is still running. If they stop receiving the heartbeat, they switch to the Candidate status.
 
 The client can ask any node to add a log to the record.
-The request includes the number of the log that the client wants to add and the log himself.
-If the client as a node which is not the leader, the request is transfered to the leader.
-If there is already a record at that number,
+A log, in my implementation, is just a string, corresponding to a unique number.
+Both are given by the client in his request to the state machine.
+If the client asks a node which is not the leader, the request is transfered to the leader. If there is already a log at the given number,
 the leader returns an error message to the client. Otherwise, the leader updates its record and asks all his followers to update themselves as well.
 
-When requesting a log, the client can also query any node in the network. The node just responds with the log at that number in the local record that it is holding.
+The client can also query any node in the network.
+The node just responds with the log at that number in the local record that it is holding.
+All nodes should have to same information and the result should be the same no matter which node the client queries.
 
-The only parameter than I included in my implementation is \emph{updateSystem}, which is a boolean. If $True$, if the leader sends heartbeats and a node does not reply, 
+The only parameter than I included in the classical consensus version of my state machine is \emph{updateSystem}, which is a boolean.
+If $True$, if the leader sends heartbeats and a node does not reply,
 it will be eliminated from the network in the knowledge of all the nodes. This makes it possible to
@@
 
-\subsection{Byzantine fault-tolerance (BFT) with classical consensus}
+% \subsection{Byzantine fault-tolerance (BFT) with classical consensus}
 
-Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way.
+% Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way.
 In practice, it means that the node will stay up and can send malicious requests to
-interfere with network usage.
+% interfere with network usage.
 
-The Raft consensus algorithm is not BFT.
+% The Raft consensus algorithm is not BFT.
 BFT algorithms (such as \cite{pbft}) include a layer of authentification, but are not necessarily much slower.
 However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
-For example, assuming partial synchrony of the nodes, consensus protocols can handle $t$ crash failures with $2t+1$ nodes but $t$ byzantine failures with no less than $3t+1$ nodes.
+% For example, assuming partial synchrony of the nodes, consensus protocols can handle $t$ crash failures with $2t+1$ nodes but $t$ byzantine failures with no less than $3t+1$ nodes.
 
@@
 
-Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. Nakamoto consensus is probabilistic, which means that there is a probability ε that consensus is not reached.
-Of course, ε can be so low that crucial systems can be built relying on Nakamoto consensus. Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
+Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. It is a probabilistic type of consensus, which means that there is a probability ε that consensus is not reached.
+Of course, ε can be so low that crucial systems can rely on the assumption that consensus is achieved.
+
+Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
 Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.
 
-Nakamoto consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state-machine with it ?
+Nakamoto consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state machine with it ?
 
@@
 
-The goal of snowball is to reach binary consensus starting from a network of nodes.
+The goal of Snowball is to reach binary consensus starting from a network of nodes.
 
-I adapted the protocol to fit the needs of a decentralized state machine, because snowball is described in the context of building a payment system (it is the basis of Avalanche, which is that payment system).
-Snowball is built by iterations, but I will only focus on the last iteration, because it is the most interesting and the one I tested the performances in the end. 
-The other two algorithms, Slush and Snowflake are only interesting to understand the protocol.
+I adapted the protocol to fit the needs of a decentralized state machine, because Snowball is described in the context of building a payment system (it is the basis for Avalanche, which is that payment system).
+Snowball is built by iterations, but I will only focus on the last iteration, because it is the most interesting and the one we will benchmark in the end. 
+The other two algorithms, Slush and Snowflake, are only for understanding purpose.
 
-I adapted Snowball in the following ways : Each node, in addition to having all the snowball consensus related data (not much), has a version of the record. All nodes starts with a empty record.
+I adapted Snowball in the following ways : Each node, in addition to having all the Snowball consensus related data (not much), has a version of the record, which is a standard key/value data structure (\href{https://go.dev/blog/maps}{a Go map}).
+All nodes starts with a empty record.
 Because it is the requirement for a decentralized state machine, each node can be queried to add a log to the record. Each node can also be queried to get one log in the record.
 
-The snowball protocol is triggered when the client asks to add a new log to an index that has not been touched yet.
-Indeed, the system works as if each index in the record corresponds to an independant snowball run.
+The Snowball protocol is triggered when the client asks to add a new log to an index that has not been used yet.
+Indeed, the system works as if each index in the record corresponds to an independant Snowball run.
 
-In snowball, each node has a counter of "confidence" which can tip towards one way or the other.
+In Snowball, each node has a counter of "confidence" ($cnt$) which can tip towards one side of the binary consensus or the other.
 In my implementation, there is no binary consensus because the only value the system has is the log request.
+Therefore, the "binary consensus" is made between the current value of the record at one position and any other value.
+
 When receiving an add-log request, the node follows this sequence, in which $logToAdd$(string) and $logPosition$(int) are variables from the request (client or other node) and $record$(map[int]string) and $cnt$(map[int]int) are variables from the node memory:
+\\
 \begin{algorithmic}
@@
     \item sampleSize (k) : the number of nodes that are queried each round
-    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successfull.
-          (with $\alpha \geq \lceil k/2 \rceil$)
-    \item CounterThreshold (β) : the threshold of successfull requests above which the node stops initiating requests and only replies with its content. (for one position)
+    \item majorityThreshold (α) : the number of nodes that have to reply successfully to consider that the query is successful.
+          (with $\alpha > \lfloor k/2 \rfloor$)
+    \item CounterThreshold (β) : the threshold of successful requests above which the node stops initiating requests and only replies with its content. (for one position)
 \end{itemize}
@@
 In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
-of values (n, k, β). (α is not included because we assume no byzantine nodes in this section).
+of values (n, k, β). (α is not included because we assume no byzantine nodes in this section, that corresponds to the implementation).
 
-The evolution of the system of nodes, for one position (one snowball run) is a Markov process. To compute to probability of not reaching consensus, we need to identify the possible states in the
+The evolution of the system of nodes, for one position (one Snowball run), is a Markov process. To compute to probability of not reaching consensus, we need to identify the possible states in the
 Markov process, the possible transitions and the transitions probabilities.
+
+Servers are represented by nodes. A server which received the request to add the log is considered "colored".
 
@@
 If the state converges to $c = n$, consensus is reached.
-If the state converges to $reqs = 0$, with $c \neq 0$, consensus is not reached when the snowball run is done.
+If the state converges to $reqs = 0$, with $c \neq 0$, consensus is not reached when the Snowball run is done.
 
@@
 That is true if and only if states are transient, which is the case here because c can only grow and with c constant, reqs can only diminish.
+This property also ensures that the system will converge towards no requests.
 
@@
 The sum represents the resolution of one group of requests of the consensus protocol for one position (t represent the number of neutral nodes touched)
+
+The starting point of the Markov process is 1 colored node and 1 group of requests (the client queries one node, which is therefore colored.
+This node sends a group of requests to the first sample of the network)
+\begin{equation*}
+    P(1, 1) = 1
+\end{equation*}
+\\
+With the starting point and the recursion formula, we can compute the values of ε. See Appendices
 
@@
 \\
-With these formulas, we can compute the probability of not reaching consensus : view Appendices.
-The scripts which were used to compute the graphs in Appendix are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}
+The starting point of the Markov process is 1 semi-colored node and 1 group of requests from a semi-colored node. 
+\begin{equation*}
+    P(0, 1, 0, 1) = 1
+\end{equation*}
+\\
+With these formulas, we can compute the probability of not reaching consensus : see Appendices.
+The scripts which were used to compute the graphs in Appendices are available at \href{https://github.com/Nicolascrd/researchProjectConsensus}{github.com/Nicolascrd/researchProjectConsensus}.
+They use Dynamic Programming.
 
@@
 \paragraph{Snowball (Nakamoto)}
-The number of requests required to reach consensus on an add-log request is $k*\beta$ for each node.
+Each node emits requests until his counter reaches $\beta$ for the position. Each time, it emits $k$ request.
 
+In total, the number of requests required to reach consensus (assuming consensus is reached) on one position is $k*\beta*n$
 \subsection{In case of a failure}
@@
 If a node crashs in Snowball, some add-log request will not find a majority and therefore the consensus will be slower and require more requests.
-However, in production, the Snowball state-machine would include a mechanism to update the network knowledge among members, which can take the form of a Snowball run, and therefore continue as if no node had failed.
+However, in production, the Snowball state machine would include a mechanism to update the network knowledge among members, which can take the form of a Snowball run, and therefore continue as if no node had failed.
 Snowball is also very efficient at managing crash failures.
@@
 \paragraph{Byzantine failures}
-The architecture required to test properly my implementations for byzantine failures goes beyong this project. But we can still think about the resilience of each implementation.
+The architecture required to test properly my implementations for byzantine failures goes beyond this project. But we can still think about the resilience of each implementation.
 With Raft, a byzantine node can easily take the control on the network by becoming the leader.
 It only needs to start as a follower, choose a currentTerm bigger than the current one and ask for votes, which other nodes will give.
-Once elected leader, the byzantine node can implement any information among followers. The state-machine is unusable and even potentially dangerously infected.
+Once elected leader, the byzantine node can implement any information among followers. The state machine is unusable and even potentially dangerously infected.
 
@@
 We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties. 
-In normal running state, the number of requests to reach consensus is a function of $k * \beta$ whereas in Raft, the leader must handle $n-1$ request on its own. 
-Raft has one consensus for the whole system, whereas in Snowball to each value correspond a different consensus. For most usage, classical consensus is probably the way to go.
-It would be useful to lead some more research on the topic. 
+In normal running state, the number of requests to reach consensus with Snowball is a function of $k * \beta$ for each node. In Raft, the leader must handle $n-1$ request on its own. 
+Raft has one consensus for the whole system, whereas in Snowball to each value correspond a different consensus.
+For most applications, classical consensus is probably the way to go.
+Still, it would be useful to lead some more research on the topic.
 In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 

### 2022-11-02 article/article.tex
typos

@@
 In Raft, each node can be either Leader, Follower or Candidate. In regular functioning, one node will be the leader and all the others will follow.
-The Leader node sends regular heartbeat requests to notify Followers that it is still running. If they stop receiving the heartbeat, they switch to the Candidate status.
+The Leader sends regular heartbeat requests to notify Followers that it is still running.
+If they stop receiving the heartbeat, they switch to the Candidate status, in which they ask other nodes to vote for them in order to become Leader.
 
@@
 
-Servers are represented by nodes. A server which received the request to add the log is considered "colored".
+Servers are represented by nodes. A server which received the request to add the log is considered "colored", otherwise it is considered "neutral".
 
@@
 The term in the sums is considered 0 if it does not make sense.
-First sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
-Second sum represents a step of the resolution of one group of requests, coming from a colored node.
+The first sum represents a step of the resolution of one group of requests, coming from a semi-colored node, with tc (ts) the number of colored (semi colored) nodes in the sample.
+The second sum represents a step of the resolution of one group of requests, coming from a colored node.
 To do the calculation, we assume that requests from semi-colored nodes resolve faster.
@@
 \paragraph{Raft (classical)}
-When the leader is up and giving instruction to all the followers, followers don't have to initiate requests.
+When the leader is up and giving instructions to all the followers, followers don't have to initiate requests.
 Therefore in normal functioning, the leader node will make $n-1$ requests every heartbeat.
@@
 \paragraph{Fail-stop in Snowball}
-If a node crashs in Snowball, some add-log request will not find a majority and therefore the consensus will be slower and require more requests.
-However, in production, the Snowball state machine would include a mechanism to update the network knowledge among members, which can take the form of a Snowball run, and therefore continue as if no node had failed.
-Snowball is also very efficient at managing crash failures.
+With no system to update knowledge among participating nodes, nodes crashes can force nodes to use more requests to reach consensus. 
+Indeed, if in the sample there are enough crashed nodes to prevent the success of the request, the node will have to start a new group of requests.
+That requires a significant number of nodes to crash nonetheless.
+In production, the Snowball state machine would include a mechanism to update the network knowledge among members, which can take the form of a Snowball run, and therefore continue as if no node had failed.
+Snowball would then be very efficient at managing crash failures.
 
@@
 In normal running state, the number of requests to reach consensus with Snowball is a function of $k * \beta$ for each node. In Raft, the leader must handle $n-1$ request on its own. 
-Raft has one consensus for the whole system, whereas in Snowball to each value correspond a different consensus.
+Raft has one consensus for the whole system, whereas in Snowball to each value corresponds a different consensus run.
 For most applications, classical consensus is probably the way to go.
@@
 In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$. 
-Also, a comparison focused on Byzantine Fault Tolerance would also be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
+A comparison focused on Byzantine Fault Tolerance would also be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
 

### 2022-11-02 article/article.tex
typos

@@
 
-In this article, we will look at a new family of probabilitic consensus algorithms, in the context of building a decentralized state machine.
+In this article, we will look at a new family of probabilistic consensus algorithms, in the context of building a decentralized state machine.
 
@@
 Both are given by the client in his request to the state machine.
-If the client asks a node which is not the leader, the request is transfered to the leader. If there is already a log at the given number,
+If the client asks a node which is not the leader, the request is transferred to the leader. If there is already a log at the given number,
 the leader returns an error message to the client. Otherwise, the leader updates its record and asks all his followers to update themselves as well.
@@
 crash a majority of nodes and still have the state machine running and usable on all the surviving nodes.
-
-% \subsection{Byzantine fault-tolerance (BFT) with classical consensus}
-
-% Contrary to Crash Fault, Byzantine Fault \cite{byzantine} is a type of failure where you consider that the node can fail in any possible way.
-In practice, it means that the node will stay up and can send malicious requests to
-% interfere with network usage.
-
-% The Raft consensus algorithm is not BFT.
-BFT algorithms (such as \cite{pbft}) include a layer of authentification, but are not necessarily much slower.
-However, there is a maximum number of malicious nodes that the algorithm can handle safely \cite{partialSynchrony}.
-% For example, assuming partial synchrony of the nodes, consensus protocols can handle $t$ crash failures with $2t+1$ nodes but $t$ byzantine failures with no less than $3t+1$ nodes.
 
@@
 The Snowball protocol is triggered when the client asks to add a new log to an index that has not been used yet.
-Indeed, the system works as if each index in the record corresponds to an independant Snowball run.
+Indeed, the system works as if each index in the record corresponds to an independent Snowball run.
 
@@
     \EndIf
+    \\
 \end{algorithmic}
-
 initiateRequest corresponds to the following sequence :
-
+\\
 \begin{algorithmic}
@@
     \EndWhile
+    \\
 \end{algorithmic}
-
 3 parameters are used in my implementation :
@@
 However, in my implementation, the client cannot either change a log at one position.
-One thing the byzantine node cannot easily do in the Snowball implementation is register different informations among multiple nodes.
-Indeed, these nodes will want to verify the information by quering other nodes to reach consensus.
+One thing the byzantine node cannot easily do in the Snowball implementation is register different data among multiple nodes.
+Indeed, these nodes will want to verify the information by querying other nodes to reach consensus.
 The system will therefore go one way or the other.
@@
 In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$.
-A comparison focused on Byzantine Fault Tolerance would also be interesting. To do that one would need to use PBFT \cite{pbft} instead of the Raft.
+A comparison focused on Byzantine Fault Tolerance would also be interesting. To do that, one would need to use PBFT \cite{pbft} instead of the Raft.
 

### 2022-11-02 biblio/biblio-rp.tex
typos biblio

--- 2022-10-03 biblio-rp.tex
@@
     Suppose also that each processor $p$ has some private value of information \(V_{p}\). [...]
-    The question is whether for given \(m, n \geqslant 0 \), it is possible to devise an algorithm based on an exchange of messages that will allow each nonfaulty processor $p$ to compute a vector of values with an element for each of the n processors, such that :
+    The question is whether for given \(m, n \geqslant 0 \), it is possible to devise an algorithm based on an exchange of messages that will allow each non-faulty processor $p$ to compute a vector of values with an element for each of the n processors, such that :
 
     \begin{itemize}
-        \item the nonfaulty processors compute exactly the same vector;
-        \item the element of this vector corresponding to a given nonfaulty processor is the private value of that processor.
+        \item the non-faulty processors compute exactly the same vector;
+        \item the element of this vector corresponding to a given non-faulty processor is the private value of that processor.
     \end{itemize}
 
-    Note that the algorithm need not reveal which processors are faulty, and that the elements of the computed vector corresponding to faulty processors may be arbitrary; it matters only that the nonfaulty processors compute exactly the same value for any given faulty processor.
-    We say that such an algorithm achieves interactive consistency, since it allows the nonfaulty processors to come to a consistent view of the values held by all the processors, including the faulty ones.
+    Note that the algorithm need not reveal which processors are faulty, and that the elements of the computed vector corresponding to faulty processors may be arbitrary; it matters only that the non-faulty processors compute exactly the same value for any given faulty processor.
+    We say that such an algorithm achieves interactive consistency, since it allows the non-faulty processors to come to a consistent view of the values held by all the processors, including the faulty ones.
 The computed vector is called an interactive consistency (i.c.) vector.
-Once interactive consistency has been achieved, each nonfaulty processor can apply an averaging or filtering function to the i.c.
+Once interactive consistency has been achieved, each non-faulty processor can apply an averaging or filtering function to the i.c.
 vector, according to the needs of the application.
-    Since each nonfaulty processor applies this function to the same vector of values, an exact agreement is necessarily reached.
+    Since each non-faulty processor applies this function to the same vector of values, an exact agreement is necessarily reached.
 
@@
 
-In the paper, the problem is decribed as so :
+In the paper, the problem is described as so :
 \subparagraph{Setting}
@@
 
-Studying the performance of consensus algorithms is not easy, as reported in \cite{analysisperf} (“Consensus protocols are typically too complex for analytical approaches to performance evalution”) and \cite{failureDetection} limits the study to the pure speed of the protocol (providing that the consensus is indeed reached).
+Studying the performance of consensus algorithms is not easy, as reported in \cite{analysisperf} (“Consensus protocols are typically too complex for analytical approaches to performance evaluation”) and \cite{failureDetection} limits the study to the pure speed of the protocol (providing that the consensus is indeed reached).
 According to Urban and Schiper \cite{comparing},
@@
 
-The Andrew benchmark \cite{scaleperf} for example is used in the PBFT\footnote{Practical Byzantine Fault-Tolerance} paper \cite{pbft} to state that their replicated file system supporting NFS\footnote{Network File System} adds very little time cost compared to a standard NFS deamon.
+The Andrew benchmark \cite{scaleperf} for example is used in the PBFT\footnote{Practical Byzantine Fault-Tolerance} paper \cite{pbft} to state that their replicated file system supporting NFS\footnote{Network File System} adds very little time cost compared to a standard NFS daemon.
 
@@
 Therefore, computing a Paxos run each time the service is queried by a client would result in a significant amount of unecessary processing, compared to a leader-based process with Paxos runs only to tackle failures.
-Most of the time, variants of Paxos would be implemented, not the original Paxos mentionned as the Basic Protocol.
+Most of the time, variants of Paxos would be implemented, not the original Paxos mentioned as the Basic Protocol.
 
@@
 
-Even if it is often refered as so, Proof-of-Work is not in itself a consensus algorithm.
+Even if it is often referred as so, Proof-of-Work is not in itself a consensus algorithm.
 Proof-of-Work was at inception a denial-of-service counter-measure \cite{hashcash}.
@@
 Without Proof-of-Work, an attacker can potentially send millions of request to a service that has to process them in what is known as a denial-of-service attack.
-If the services requieres the request to contains a Proof-of-Work stamp, the attacker would need for example one second to compute each stamp, which would need a lot of power and time to do the attack, in practice avoiding this kind of attacks completely.
+If the services requires the request to contain a Proof-of-Work stamp, the attacker would need for example one second to compute each stamp, which would need a lot of power and time to do the attack, in practice avoiding this kind of attacks completely.
 
 In the bitcoin blockchain \cite{bitcoin}, Proof-of-Work is used to build the consensus algorithm which consists in giving trust to the longest chain\footnote{Longest chain of blocks in the blockchain, see \cite{blockchain} or \cite{bitcoin} for details}.
-This is some times referred to as the Nakamoto consensus, like in \cite{proconsensus}.
+This is sometimes referred to as the Nakamoto consensus, like in \cite{proconsensus}.
 The longest chain can only be built if you have the majority of computing power on the network.
@@
 The second biggest cryptocurrency by market cap\footnote{Market capitalization (or market cap) is the total value of all the coins that have been mined.
-It is calculated by multiplying the number of coins in circulation by the current market price of a single coin.}, Ethereum, even moved from Proof-of-Work to Proof-of-Stake not long before these lignes are written, in on of the biggest event the world of crypto has ever seen (known as The Merge).
+It is calculated by multiplying the number of coins in circulation by the current market price of a single coin.}, Ethereum, even moved from Proof-of-Work to Proof-of-Stake not long before these lines were written, in one of the biggest event the world of crypto has ever seen (known as The Merge).
 
-Proof-of-Stake consists in replacing the trust from computing power with the trust from holding the coins of the ecosystem, therefore prooving that someone has something to loose beeing dishonest.
+Proof-of-Stake consists in replacing the trust from computing power with the trust from holding the coins of the ecosystem, therefore proving that someone has something to loose by being dishonest.
 

### 2022-11-16 article/article.tex
proofreading fixes

@@
 
-\title{Decentralized state machine using Nakamoto (probabilistic) consensus\\\medskip Research Project 2022}
+\title{Decentralized State Machine using Nakamoto (probabilistic) Consensus\\\medskip Research Project 2022}
 \author{Nicolas COURAUD (nicolas.couraud@etu.emse.fr)\\École Nationale Supérieure des Mines de Saint-Étienne}
@@
 
-In this article, we will look at a new family of probabilistic consensus algorithms, in the context of building a decentralized state machine.
+In this article, we will look at a new family of probabilistic Consensus algorithms, in the context of building a decentralized state machine.
 
 We will show that it is possible to build a decentralized state machine using Nakamoto Consensus, with the Snowball protocol \cite{snowprotocol} and will compare this approach to the traditional approach with
-a classical consensus algorithm (Raft \cite{understandable})
+a classical Consensus algorithm (Raft \cite{understandable})
 
@@
 
-\section{Traditional state machine replication}
+\section{Traditional State Machine Replication}
 
@@
 
-Classical consensus algorithms were designed in the late 20th century to tackle faulty hardware and network. Their goal is to avoid one service being made unavailable by a crash or a network failure.
-In classical consensus, all of the nodes have to know each other. It is a permissioned model, which means that nodes have to get an approval to get into the network.
+Classical Consensus algorithms were designed in the late 20th century to tackle faulty hardware and network. Their goal is to avoid one service being made unavailable by a crash or a network failure.
+In classical Consensus, all of the nodes have to know each other. It is a permissioned model, which means that nodes have to get an approval to get into the network.
 Because all nodes must communicate with each other, the communication cost to reach consensus in generally considered high.
@@
 
-Many classical consensus algorithms exists, the most famous one being Paxos \cite{parliament}.
+Many classical consensus algorithms exist, the most famous one being Paxos \cite{parliament}.
 \subsection{The Raft State Machine}
@@
 
-\section{Nakamoto (probabilistic) state machine replication}
+\section{Nakamoto (probabilistic) State Machine Replication}
 \subsection{The Nakamoto Consensus}
 
-Nakamoto consensus was introduced with the bitcoin blockchain \cite{bitcoin}. It is a probabilistic type of consensus, which means that there is a probability ε that consensus is not reached.
+Nakamoto Consensus was introduced with the bitcoin blockchain \cite{bitcoin}. It is a probabilistic type of consensus, which means that there is a probability ε that consensus is not reached.
 Of course, ε can be so low that crucial systems can rely on the assumption that consensus is achieved.
 
-Nakamoto consensus is open : nodes don't have to register or get an authorisation to enter the network.
+Nakamoto Consensus is open : nodes don't have to register or get an authorisation to enter the network.
 Because nodes don't have to know every other node in the network and be able to communicate with everyone, Nakamoto is supposed to allow for greater scale in the network.
 
-Nakamoto consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state machine with it ?
+Nakamoto Consensus was a revolution in the consensus space with cryptocurrencies. How can we adapt the approach to build a state machine with it ?
 
-\subsection{My Nakamoto state machine implementation}
+\subsection{My Nakamoto State Machine Implementation}
 
-To implement Nakamoto consensus, I chose the Snowball consensus algorithm \cite{snowprotocol}, which underlies the Avalanche blockchain.
+To implement Nakamoto Consensus, I chose the Snowball consensus algorithm \cite{snowprotocol}, which underlies the Avalanche blockchain.
 
@@
 
-In order to compare the performances of classical consensus and Nakamoto consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
+In order to compare the performances of classical consensus and Nakamoto Consensus, we need to be able to know the probability of not reaching consensus ε corresponding to the set
 of values (n, k, β). (α is not included because we assume no byzantine nodes in this section, that corresponds to the implementation).
@@
 
-\paragraph{Raft (classical)}
+\paragraph{Raft (Classical)}
 When the leader is up and giving instructions to all the followers, followers don't have to initiate requests.
@@
 However, the number of requests that will be made to elect a new leader depends on a number of factors.
-With a few nodes, the process happens with only one request : the first node to tick (notice that the heartbeat was not received) switches to candidate status, triggers a round of vote, collects all the votes and establishes itself as the new leader.
-In the meantime, the other nodes did not switch to candidate and start receiving the heartbeats from the new leader.
+With a few nodes, the process happens with only one request : the first node to tick (notice that the heartbeat was not received) switches to candidate status, triggers a round of voting, collects all the votes and establishes itself as the new leader.
+In the meantime, the other nodes have not switched to candidate and start receiving the heartbeats from the new leader.
 With many nodes, nodes might tick "at the same time" and therefore candidate at the same time which can trigger many requests.
@@
 \paragraph{Fail-stop in Snowball}
-With no system to update knowledge among participating nodes, nodes crashes can force nodes to use more requests to reach consensus.
+With no system to update knowledge among participating nodes, node crashes can force nodes to use more requests to reach consensus.
 Indeed, if in the sample there are enough crashed nodes to prevent the success of the request, the node will have to start a new group of requests.
@@
 \paragraph{Byzantine failures}
-The architecture required to test properly my implementations for byzantine failures goes beyond this project. But we can still think about the resilience of each implementation.
-With Raft, a byzantine node can easily take the control on the network by becoming the leader.
+The architecture required to benchmark my implementations for byzantine failures goes beyond this project. But we can still think about the resilience of each implementation.
+With Raft, a byzantine node can easily take control on the network by becoming the leader.
 It only needs to start as a follower, choose a currentTerm bigger than the current one and ask for votes, which other nodes will give.
@@
 This would require the consensus to switch from one way to the other, which is nearly impossible.
-However, in my implementation, the client cannot either change a log at one position.
+However, in my implementation, the client cannot change a log at one position either.
 One thing the byzantine node cannot easily do in the Snowball implementation is register different data among multiple nodes.
@@
 
-We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto consensus algorithm and had interesting properties.
+We demonstrated that it was possible to build a decentralized state machine, that could be based on a Nakamoto Consensus algorithm and had interesting properties.
 In normal running state, the number of requests to reach consensus with Snowball is a function of $k * \beta$ for each node. In Raft, the leader must handle $n-1$ request on its own.
 Raft has one consensus for the whole system, whereas in Snowball to each value corresponds a different consensus run.
-For most applications, classical consensus is probably the way to go.
-Still, it would be useful to lead some more research on the topic.
+For most applications, Classical Consensus is probably the way to go.
+Still, it would be useful to conduct some more research on the topic.
 In particular, it would be interesting to approximate the value of ε $\forall n, k, \beta$.
